---
title: "Predicting Absenteeism at work"
subtitle: "Machine Learning II"
author: "Sandra Alemayehu & Jessica Hayek"
date: "2/21/2020"
output: html_document

---

```{r, message=FALSE, warning=FALSE, include=FALSE}
library(MASS)
library(ISLR)
library(rpart)
library(rpart.plot)
library(randomForest)
library(ggplot2)
library(e1071)     
library(glmnet)
library(caret)     
library(FSelector)
library(dplyr)
library(knitr)
library(magrittr)
library(lattice)
```

#Introduction


#Loading Data

The dataframe contains **22** variables.The target variable is **Absenteeism**, a logical variable setting categorizing whether an individual is going to be absent more than 5 hours in total.

```{r Loading Training}

abs_training_df<-read.csv("Absenteeism_at_work_classification_training.csv",sep = ";")
ncol(abs_training_df)
```


```{r Variables}
data.frame(Variable = names(abs_training_df),
           Class = sapply(abs_training_df, typeof),
           First_values = sapply(abs_training_df, function(x) paste0(head(x),  collapse = ", ")),
           row.names = NULL) %>% kable()

```



```{r cars}

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
